---
title: "04_Noah"
author: '38880242'
date: "2025-06-05"
output: html_document
---

#############
### SETUP ###
#############

# install.packages(c("ggplot2", "tidyverse"))
library(ggplot2)
library(tidyverse)

##############
### PART 1 ###
##############


# load data
```{r}
field_goals = read_csv("../data/04_field-goals.csv.gz")

linearregressionmodel <- lm(fg_made ~ ydl + kq, data = field_goals)
logit_model <- glm(fg_made ~ ydl + kq, data = field_goals)
spline_lpm <- lm(fg_made ~ bs(ydl, df = 4) + kq,data   = field_goals)

summary(logit_model)
summary(linearregressionmodel)
summary(spline_lpm)
```
```{r}

```

```{r}

set.seed(123)
n_total   <- nrow(field_goals)
train_n   <- floor(0.7 * n_total)
train_ix  <- sample(1:n_total, train_n, replace = FALSE)

train_data <- field_goals[train_ix, ]
test_data  <- field_goals[-train_ix, ]


linearregressionmodel <- lm(
  fg_made ~ ydl + kq,
  data   = train_data      
)

logit_model <- glm(
  fg_made ~ ydl + kq,
  data   = train_data    
)

spline_lpm <- lm(
  fg_made ~ bs(ydl, df = 2) + kq,
  data   = train_data      
)

test_data$pred_lpm <- predict(linearregressionmodel, newdata = test_data, type = "response")
test_data$pred_logit  <- predict(logit_model,        newdata = test_data, type = "response")
test_data$pred_spline <- predict(spline_lpm,          newdata = test_data, type = "response")

mse_lpm <- mean((test_data$pred_lpm    - y_true)^2)
mse_logit <- mean((test_data$pred_logit  - y_true)^2)
mse_spline <- mean((test_data$pred_spline - y_true)^2)

```

The intercept being .974 means that when both `ydl` and `kq` are zero, the model predicts a field goal made rate of above 1 

Since these are splines, the coefficients of -.1119 means that holding kicker quality constant, as you move up towards the knot then the predicted field goal made rate decreases by .1119 for each unit increase in `ydl`.
```{r}
summary(spline_lpm)
```

```{r}
ggplot(test_data, aes(x = pred_spline, color = factor(fg_made), fill = factor(fg_made))) +
  geom_density() +
  scale_x_continuous(
    name   = "Predicted Probability",
    limits = c(0, 1)
  ) +
  labs(
    color = "Actual Outcome",
    fill  = "Actual Outcome"
  ) +
  theme_minimal()


```

##############
### PART 2 ###
##############

# load data

```{r}
ncaab_results = read_csv("../data/04_ncaab-results.csv.gz")
ncaab_team_info = read_csv("../data/04_ncaab-teams.csv.gz")

ncaab_23 <- ncaab_results %>%
  filter(Season == 2023)

game_bt <- ncaab_23 %>%
  mutate(
    HomeTeam = if_else(WLoc == "A", LTeamID, WTeamID),
    AwayTeam = if_else(WLoc == "A", WTeamID, LTeamID),
    y        = if_else(WLoc == "A", 0, 1),
    beta0    = if_else(WLoc == "N", 0.5, 0)
  ) %>%
  select(HomeTeam, AwayTeam, y, beta0) %>%
  mutate(
    HomeTeam = factor(HomeTeam),
    AwayTeam = factor(AwayTeam)
  )

all_teams <- sort(
  unique(
    c(
      as.character(game_bt$HomeTeam),
      as.character(game_bt$AwayTeam)
    )
  )
)

K <- length(all_teams)
n <- nrow(game_bt)

home_idx <- match(as.character(game_bt$HomeTeam), all_teams)
away_idx <- match(as.character(game_bt$AwayTeam), all_teams)


# 2A. Initialize an all‐zero matrix of dimension n × K
X <- matrix(0, nrow = n, ncol = K)

# 2B. Create a vector of row‐indices 1, 2, …, n
row_ids <- seq_len(n)

# 2C. Place +1 in (row_i, home_idx[i]) for every i
X[cbind(row_ids, home_idx)] <- +1

# 2D. Place –1 in (row_i, away_idx[i]) for every i
X[cbind(row_ids, away_idx)] <- -1


```

```{r}
y_vec     <- game_bt$y      
beta0_vec <- game_bt$beta0  


df_design <- as.data.frame(X)


colnames(df_design) <- all_teams

df_design$beta0 <- beta0_vec
df_design$y     <- y_vec

bt_fit <- glm(
  y ~ beta0 + . - 1,
  data   = df_design,
  family = binomial(link = "logit")
)

summary(bt_fit)
```

```{r}
team_coefs <- coef(bt_fit)[ names(coef(bt_fit)) != "beta0" ]


df_coefs <- data.frame(
  team     = names(team_coefs),
  strength = as.numeric(team_coefs),
  stringsAsFactors = FALSE
) %>%
  arrange(desc(strength)) %>%
  mutate(team = factor(team, levels = team))

library(ggplot2)
ggplot(df_coefs, aes(x = team, y = strength)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs() +
  theme_minimal()


```




