4.1.2
```{r}
#1 - 3 different models for fg success prob 

library(ggplot2)
library(tidyverse)

fg = read_csv('../data/04_field-goals.csv.gz')
head(fg)

plotting <- function(x_data, y_data, color, x_title, y_title) {
  ggplot(punting, aes(x = x_data, y = y_data, color = color)) +
  geom_point(alpha = 0.7, size = 2) +
  labs(
      x = x_data,
      y = y_data,
      color = color
    ) +
    theme_minimal()

}

#plotting(fg$ydl, fg$fg_made, 'cornflowerblue', "Field Goal Yard Lines", "Kicker Quality") +
  #geom_smooth(method = 'lm', se = FALSE, color = 'black') +
  #ggtitle("Field Goal Success Probability by Yard Line")

#linear model
model = lm(fg_made ~ ydl + kq, data = fg)
model
#logistic model
log_model = glm(fg_made ~ ydl + kq, data = fg, family = binomial)
log_model
#logistic polynomial model 
log_poly_model = glm(fg_made ~ poly(ydl, 3) + kq, data = fg, family = binomial)
log_poly_model
```

```{r}
#2 - out of sample performance of each model 
#performance function from lab 3
performance <- function(data, formula, model_func = lm, test_frac = 0.2, seed = 42, family = NULL) {
  set.seed(seed)
  
  n <- nrow(data)
  test_idx <- sample(1:n, size = test_frac * n)
  
  train <- data[-test_idx, ]
  test <- data[test_idx, ]
  
  # Fit model with user-defined modeling function and family
  if (!is.null(family)) {
    model <- model_func(formula = formula, data = train, family = family)
  } else {
    model <- model_func(formula = formula, data = train)
  }
  
  # Predict
  test$pred <- predict(model, newdata = test, type = if ("glm" %in% class(model)) "response" else "response")
  
  # Evaluate
  response_var <- all.vars(formula)[1]
  actual <- test[[response_var]]
  mse <- mean((actual - test$pred)^2)
  r2 <- 1 - sum((actual - test$pred)^2) / sum((actual - mean(actual))^2)
  
  return(list(
    formula = formula,
    model_type = deparse(substitute(model_func)),
    mse = mse,
    r_squared = r2
  ))
}

#evaluate the models
performance(fg, fg_made ~ ydl + kq, model_func = lm)
performance(fg, fg_made ~ ydl + kq, model_func = glm)
performance(fg, fg_made ~ poly(ydl, 3) + kq, model_func = glm)
```
3 - Interpretation of coefficients: 
We see that both the linear model and the first logistic model are not the most accurate. The polynomial logistic model offers a better mse, meaning that the prediction is about .001 squared units more accurate, or .01 more units more accurate. This means that the model is more accurate on ~1/100 field goals. 

```{r}
#4 - plot the predicted probabilities of success for best model

#actual probabilities based on yard line
grouped_ydl <- fg %>%
  group_by(ydl) %>%
  summarise(avg_kq = mean(kq), avg_fg_made = mean(fg_made))

grouped_ydl <- grouped_ydl %>%
  rename(kq = avg_kq)

#predicted probabilities based on yard line 
x = seq(min(fg$ydl), max(fg$ydl), length.out = 100)
predicted_prob <- predict(log_poly_model, newdata = data.frame(ydl = x, kq = mean(fg$kq)), type = "response")
grouped_ydl <- grouped_ydl %>%
  mutate(predicted_prob = predict(log_poly_model, newdata = data.frame(ydl = ydl, kq = mean(fg$kq)), type = "response"))

head(grouped_ydl)

#plot data with prediction
ggplot(grouped_ydl, aes(x = ydl, y = avg_fg_made)) +
  geom_point(aes(color = kq), size = 2, alpha = 0.7) +
  geom_line(aes( y = predicted_prob), color = "black") +
  labs(
    x = "Field Goal Yard Lines",
    y = "Average Field Goal Success Probability",
    color = "Average Kicker Quality"
  ) +
  theme_minimal() +
  ggtitle("Predicted Probabilities of Field Goal Success (Fitted Model)")
```
4.2 - bradley-terry ncaa men's basketball scores

```{r}
#1 - filter data 

ncaab_results = read_csv("../data/04_ncaab-results.csv.gz")
ncaab_team_info = read_csv("../data/04_ncaab-teams.csv.gz")

head(ncaab_results)
head(ncaab_team_info)

ncaab_results <- ncaab_results %>%
  filter(Season == 2023) %>% 
  mutate(home_win = ifelse(WLoc == 'H', 1, 0)) 

head(ncaab_results)

```

```{r}
#2 - bradley terry model
ncaab_named <- ncaab_results %>%
  left_join(ncaab_team_info, by = c("WTeamID" = "TeamID")) %>%
  rename(WTeamName = TeamName)

# Merge loser names
ncaab_named <- ncaab_named %>%
  left_join(ncaab_team_info, by = c("LTeamID" = "TeamID")) %>%
  rename(LTeamName = TeamName)

all_teams <- sort(unique(c(ncaab_named$WTeamName, ncaab_named$LTeamName)))

# Create the matrix
home_away_matrix <- matrix(0, nrow = nrow(ncaab_named), ncol = length(all_teams))
colnames(home_away_matrix) <- all_teams

for (i in seq_len(nrow(ncaab_named))) {
  w <- as.character(ncaab_named$WTeamName[i])
  l <- as.character(ncaab_named$LTeamName[i])
  loc <- ncaab_results$WLoc[i]
  
  if (loc == "H") {
    home_away_matrix[i, w] <- 1
    home_away_matrix[i, l] <- -1
  } else if (loc == "A") {
    home_away_matrix[i, w] <- -1
    home_away_matrix[i, l] <- 1
  } else if (loc == "N") {
    home_away_matrix[i, w] <- .5
    home_away_matrix[i, l] <- .5
  }
}

home_away_matrix <- cbind(
  home_advantage = ifelse(ncaab_results$WLoc == "N", 0.5, 1),
  home_away_matrix
)

nrow(home_away_matrix)
nrow(ncaab_named)

home_away_df <- as.data.frame(home_away_matrix)
home_away_df$home_advantage <- ifelse(ncaab_results$WLoc == "N", 0.5, 1)

bradleyTerry = glm(ncaab_results$home_win ~ home_away_matrix , family = binomial())
bradleyTerry

```

```{r}
# Extract coefficients and exponentiate

coeffs <- coef(bradleyTerry)
team_coeffs <- coeffs[!names(coeffs) %in% "home_advantage"] 

team_coeffs <- coeffs[!names(coeffs) %in% c("(Intercept)", "home_advantage")]
clean_names <- gsub("^home_away_matrix", "", names(coeffs))

keep_indices <- !(clean_names %in% c("home_advantage", "(Intercept)"))

team_coeffs <- coeffs[keep_indices]
team_names  <- clean_names[keep_indices]

# 3. Create cleaned data frame
coeff_df <- data.frame(
  Team = team_names,
  Strength = team_coeffs 
)

head(coeff_df)

library(ggplot2)

ggplot(coeff_df, aes(x = reorder(Team, Strength), y = Strength)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(
    title = "Estimated Team Strengths (Bradley-Terry Model)",
    x = "Team",
    y = "Strength (Odds Multiplier)"
  ) +
  theme_minimal()

```
```{r}
#4 - vegas odds

purdue = coeff_df %>% 
  filter(Team == "Purdue") %>%
  pull(Strength)

purdue 
uconn = coeff_df %>% 
  filter(Team == "Connecticut") %>%
  pull(Strength)

uconn 


p = 1/(1+exp(-(purdue-uconn)))
p

```

