---
title: "Lab 7"
output:
  pdf_document: default
  html_document: default
date: "2025-06-10"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

Diving

```{r dive, echo = FALSE, message = FALSE}
library(tidyverse)
library(readr)
library(purrr)
library(broom)
diving_data = read_csv("../data/07_diving.csv.gz")

data = diving_data %>%
  group_by(Diver, DiveNo, Round, Event) %>%
  mutate(avg_score = mean(JScore, na.rm = TRUE)) %>% 
  ungroup() %>% 
  mutate(disc = JScore - avg_score)


# make a permutatin test for each judge comparing the disc for competitors that match country vs those that do not

Judge = unique(data$Judge)

p_values = c()
judge_dods =c()
set.seed(123)

for (judge in Judge) {
  judge_data = data %>% filter(Judge == judge)
  
    # Get the competitors from the same country
    same_country = judge_data %>% filter(Country == JCountry)
    same_disc = same_country$disc
    
    # Get the competitors from different countries
    different_country = judge_data %>% filter(Country != JCountry)
    diff_disc = different_country$disc
    
    judge_dod = mean(same_disc) - mean(diff_disc)
    judge_dods = c(judge_dods, judge_dod)
    
    perm_dod = c()
    
    for (i in 1:1000) {
      # Shuffle the countries 
      judge_data$Country = sample(judge_data$Country)
      
      # Get the competitors from the same country
      same_country = judge_data %>% filter(Country == JCountry)
      same_disc = same_country$disc
    
      # Get the competitors from different countries
      different_country = judge_data %>% filter(Country != JCountry)
      diff_disc = different_country$disc
      
      # Calculate the difference in means
      perm_dod[i] = mean(same_disc) - mean(diff_disc)
    }
    
    pvalue = mean(perm_dod >= judge_dod)
    p_values = c(p_values, pvalue)
   
    
  
}
results = data.frame(Judge = Judge, p_value = p_values, judge_dod = judge_dods)
results %>% 
  arrange(p_value)
```
Alt, Mena , Zaitsev, and Barnett all have a pvalue of 0.000, suggesting strong bias. others also have very low pvalues, but those are the worst


Beisbol

```{r pressure, echo=FALSE}
mlb_data = read.csv("../data/07_tto.csv.gz")
model1 = lm(EVENT_WOBA_19 ~ factor(ORDER_CT) + WOBA_FINAL_BAT_19 + WOBA_FINAL_PIT_19 + factor(HAND_MATCH) + factor(BAT_HOME_IND), data = mlb_data)

model2 = lm(EVENT_WOBA_19 ~ factor(ORDER_CT) + WOBA_FINAL_BAT_19 + WOBA_FINAL_PIT_19 + factor(HAND_MATCH) + factor(BAT_HOME_IND) + BATTER_SEQ_NUM, data = mlb_data)

```

```{r, message=FALSE, echo = FALSE}
summary(model1)
summary(model2)

```
in model one, all p values besides hand_match 0.5 are significant. however, in model 2, 
neither order_ct variables are significant at all any more. batter_seq_num is significant, implying that that is what is actually significant, with the count variables acting as a proxy for that in model1.
```{r}
```

