
7.1.2 Task

```{r}
library(ggplot2)
library(tidyverse)
#1 - permutation test for diving data 
diving = read.csv("../data/07_diving.csv.gz")
head(diving)

judges = unique(diving$Judge)
results <- data.frame(Judge = character(), p_value = numeric(), stringsAsFactors = FALSE)

n_permutations = 1000

diving_means <- diving %>%
    group_by(Diver, DiveNo, Round, Event) %>%
    summarise(mean_score = mean(JScore, na.rm = TRUE), .groups = 'drop')

head(diving_means)

#loop through each judge
for (judge in judges){
  diving_judges <- diving %>%
    filter(Judge == judge) %>%
    select(JScore, JCountry, Country, Judge, Diver, DiveNo, Round, Event) %>% 
    mutate(match = (Country == JCountry))
  
  
  judging_with_disc <- diving_judges %>%
    left_join(diving_means, by = c("Diver", "DiveNo", "Round", "Event")) %>%
    mutate(discrepancy = JScore - mean_score)
  
  observed_diff = mean(judging_with_disc$discrepancy[judging_with_disc$match], na.rm = TRUE) -
       mean(judging_with_disc$discrepancy[!judging_with_disc$match], na.rm = TRUE)
  print(observed_diff)
  
  perm_diffs <- numeric(n_permutations)
  for (i in 1:n_permutations) {
    perm_country <- sample(judging_with_disc$Country)  # permute diver nationalities
    
    judging_perm <- judging_with_disc %>%
      mutate(perm_match = perm_country == JCountry)
    
    perm_diffs[i] <- mean(judging_perm$discrepancy[judging_perm$perm_match], na.rm = TRUE) -
                     mean(judging_perm$discrepancy[!judging_perm$perm_match], na.rm = TRUE)
  }

  # 6. Compute p-value (two-sided)
  p_val <- mean(abs(perm_diffs) >= abs(observed_diff))

  # 7. Save result
  results <- rbind(results, data.frame(Judge = judge, observed_diff = observed_diff, p_value = p_val))
}

head(results)
results <- results %>% 
  arrange(desc(observed_diff))

head(results)

```

Michael Geissbuhler, Hans-Peter Burk, Walter ALT, and Peter HUBER have the highest discrepancies and all of which are statistically significant


```{r}
```

7.2.2. Task 

```{r}
#fit model 1 from 7.2.1
tto = read.csv("../data/07_tto.csv.gz")
head(tto)


tto_model_1 <- lm(EVENT_WOBA_19 ~ factor(ORDER_CT)+ WOBA_FINAL_BAT_19+WOBA_FINAL_PIT_19+factor(HAND_MATCH)+factor(BAT_HOME_IND) + 0, data = tto)
summary(tto_model_1)
```
We see that all the p-values are below 0 except for a hand match, which is likely due to the fact that there is a low amount of hand match values 0.5. Thus, these coefficients are likely reliable.
```{r}
#fit model 2 
tto_model_2 <- lm(EVENT_WOBA_19 ~ factor(ORDER_CT) + BATTER_SEQ_NUM + WOBA_FINAL_BAT_19+WOBA_FINAL_PIT_19+factor(HAND_MATCH)+factor(BAT_HOME_IND), data = tto)
summary(tto_model_2)

```
We see here that the p values drastically increase for the times through the order terms. This is likely because the batting sequence number is much more important to the regression model and thus the TTO terms are not as important. 

Thus, we can conclude that pitcher decline through TTO is not significant since p >> 0.05. Pitcher fatigue as a game progresses (represented by the batter sequence number) is likely a greater factor in the regression model than TTO.
