---
title: "Lab 8"
output: html_notebook
---

8.1.2 Task

```{r}
#1 - filter to include only at least 25 FT attempted 

library(ggplot2)
library(tidyverse)
set.seed(8)

nba = read.csv('../data/08_nba-free-throws.csv.gz', sep = ';')
head(nba)


nba$FTM <- nba$FT. * nba$FTA 
nba$tot_FTA <- nba$FTA * nba$G
nba$tot_FTM <- nba$FTM * nba$G

nba_cleaned <- nba %>% 
  group_by(Player) %>%
  summarise(
    FTA = sum(FTA, na.rm = TRUE),
    tot_FTA = sum(tot_FTA, na.rm = TRUE),
    tot_FTM = sum(tot_FTM, na.rm = TRUE),
    FT_percent = sum(FT., na.rm = TRUE) / n(),  # or compute FT% from FTM/FTA if you have that
    .groups = 'drop'
  ) %>%
  filter(tot_FTA >= 25) %>% 
  select(Player, FTA, tot_FTA, FT_percent, tot_FTM)

head(nba_cleaned)

nba_summary <- nba_cleaned %>%
  mutate(
    wald_se = sqrt((FT_percent * (1 - FT_percent)) / FTA),
    wald_lower = FT_percent - 1.96 * wald_se,
    wald_upper = FT_percent + 1.96 * wald_se
  )

nba_summary <- nba_summary %>%
  mutate(
    n_tilde = tot_FTA + 4,
    p_tilde = (tot_FTM + 2) / n_tilde,
    ac_se = sqrt((p_tilde * (1 - p_tilde)) / n_tilde),
    ac_lower = p_tilde - 1.96 * ac_se,
    ac_upper = p_tilde + 1.96 * ac_se
  )

nba_summary <- nba_summary %>%
  slice_sample( n = 25)


ggplot(nba_summary, aes(x = reorder(Player, -FT_percent), y = FT_percent)) +
  geom_bar(stat = "identity", fill = "orchid", color = "black") +
  geom_errorbar(aes(ymin = wald_lower, ymax = wald_upper), width = 0.2, color = "darkmagenta", alpha = 0.7) +
  geom_errorbar(aes(ymin = ac_lower, ymax = ac_upper), width = 0.2, color = "steelblue", linetype = "dashed", alpha = 0.7) +
  coord_flip() +
  labs(
    title = "Free Throw % with Wald and Agresti-Coull Intervals (≥ 25 Attempts)",
    x = "Player",
    y = "Free Throw Percentage"
  ) +
  theme_minimal()


```
We can see that the Agresti - Coull intervals are generally much lower than the Wald intervals, showing how they are much more stable 

Task 8.2.1

```{r}
#1 - create a sequence of 1000 values between 0 and 1
interval = seq(0, 1, length.out = 1000)

# 2 - for each n and p generate free throws 
n_values = c(10, 50, 100, 250, 500, 1000)
reps <- 1:100
sim_grid <- expand.grid(p = interval, n = n_values, rep = reps)
sim_grid$made_fts <- mapply(function(p, n) rbinom(1, n, p), sim_grid$p, sim_grid$n)
sim_grid$FT_percent <- sim_grid$made_fts / sim_grid$n



#3 compute confidence intervals
  simulation <- sim_grid %>% 
  mutate(
    wald_se = sqrt(FT_percent*(1-FT_percent)/n),
    wald_lower = FT_percent - 1.96*wald_se ,
    wald_upper = FT_percent + 1.96*wald_se ,
    n_tilde = n + 4 ,
    p_tilde = (made_fts + 2) / n_tilde ,
    ac_se = sqrt((p_tilde * (1 - p_tilde)) / n_tilde),
    ac_lower = p_tilde - 1.96 * ac_se,
    ac_upper = p_tilde + 1.96 * ac_se
  )


#5 plot coverage
coverage <- simulation %>%
  mutate(
    wald_covered = (p >= wald_lower & p <= wald_upper),
    ac_covered = (p >= ac_lower & p <= ac_upper)
  ) %>%
  group_by(p,n) %>%
  summarise(
    wald_coverage = mean(wald_covered),
    ac_coverage = mean(ac_covered),
    .groups = 'drop'
  )

coverage_long <- coverage %>%
  pivot_longer(cols = c(wald_coverage, ac_coverage),
               names_to = "method", values_to = "coverage")

ggplot(coverage_long, aes(x = p, y = coverage, color = as.factor(n))) +
  geom_line(size = 1) +
  geom_hline(yintercept = 0.95, linetype = "dashed", color = "gray") +
  facet_wrap(~ method, ncol = 1) +
  labs(
    title = "Empirical Coverage of Confidence Intervals",
    x = "True Probability (p)",
    y = "Coverage",
    color = "Sample Size (n)"
  ) +
  theme_minimal()

```

9.2.1 Task

```{r}
#1 - bootstrap for true FT% of all NBA players
set.seed(42)
n_boot <- 1000

# Bootstrap function for a single player
bootstrap_ci <- function(makes, total, reps = 1000) {
  boot_props <- replicate(reps, {
    sample <- sample(c(1, 0), size = total, replace = TRUE, prob = c(makes / total, 1 - makes / total))
    mean(sample)
  })
  quantile(boot_props, probs = c(0.025, 0.975))
}

# Apply per player
boot_intervals <- nba_cleaned %>%
  mutate(
    boot_ci = map2(tot_FTM, tot_FTA, ~bootstrap_ci(.x, .y, n_boot)),
    boot_lower = map_dbl(boot_ci, 1),
    boot_upper = map_dbl(boot_ci, 2)
  ) %>%
  select(-boot_ci)

```

```{r}
nba_summary <- nba_summary %>%
  left_join(boot_intervals, by = "Player")


ggplot(nba_summary, aes(x = reorder(Player, -FT_percent.x), y = FT_percent.x)) +
  geom_bar(stat = "identity", fill = "orchid", color = "black") +
  geom_errorbar(aes(ymin = wald_lower, ymax = wald_upper), width = 0.2, color = "darkmagenta", alpha = 0.7) +
  geom_errorbar(aes(ymin = ac_lower, ymax = ac_upper), width = 0.2, color = "steelblue", linetype = "dashed", alpha = 0.7) +
  geom_errorbar(aes(ymin = boot_lower, ymax = boot_upper), width = 0.2, color = "forestgreen", linetype = "dotdash", alpha = 0.9) +
  coord_flip() +
  labs(
    title = "Free Throw % with Wald, Agresti-Coull, and Bootstrap Intervals (≥ 25 Attempts)",
    x = "Player",
    y = "Free Throw Percentage"
  ) +
  theme_minimal()

```

Task 3 - Bootstrap for simulation conducted in Task 8.2.2

```{r}
n_boot = 1000

interval = seq(0, 1, length.out = 100)

# 2 - for each n and p generate free throws 
n_values = c(10, 100, 500)
reps <- 1:100
sim_grid <- expand.grid(p = interval, n = n_values, rep = reps)
sim_grid$made_fts <- mapply(function(p, n) rbinom(1, n, p), sim_grid$p, sim_grid$n)
sim_grid$FT_percent <- sim_grid$made_fts / sim_grid$n

bootstrap_ci <- function(makes, total, n_boot = 1000) {
  if (total == 0) return(c(NA, NA))  # skip rows with no attempts

  prob <- makes / total

  boot_props <- replicate(n_boot, {
    sample_attempts <- sample(c(1, 0), size = total, replace = TRUE, prob = c(prob, 1 - prob))
    mean(sample_attempts)
  })

  quantile(boot_props, probs = c(0.025, 0.975))
}

library(purrr)
sim_grid2 <- sim_grid %>%
  mutate(
    boot_ci = pmap(list(made_fts, n), bootstrap_ci, n_boot = n_boot),
    boot_lower = map_dbl(boot_ci, 1),
    boot_upper = map_dbl(boot_ci, 2)
  )

```

```{r}
sim_grid2 %>%
  filter(n == 100) %>%
  ggplot(aes(x = p, y = FT_percent)) +
  geom_point(alpha = 0.5, color = "black") +
  geom_errorbar(aes(ymin = boot_lower, ymax = boot_upper), width = 0.01, color = "steelblue", alpha = 0.6) +
  labs(
    title = "Bootstrap 95% Confidence Intervals for Observed FT%",
    subtitle = "Sample size n = 100",
    x = "True FT% (p)",
    y = "Observed FT% ± CI"
  ) +
  theme_minimal()
```

4 - coverage for the bootstrap method

```{r}

sim_grid2 <- sim_grid2 %>%
  left_join(
    simulation %>% select(p, n, wald_lower, wald_upper, ac_lower, ac_upper),
    by = c("p", "n")
  )

coverage2 <- sim_grid2 %>%
  mutate(
    bootstrap_covered = (p >= boot_lower & p <= boot_upper),
  ) %>%
  group_by(p, n) %>%
  summarise(
    bootstrap_coverage = mean(bootstrap_covered, na.rm = TRUE),
    .groups = 'drop'
  )

bootstrap_coverage <- coverage2 %>%
  pivot_longer(cols = ends_with("coverage"),
               names_to = "method", values_to = "coverage")
```
Plot these results

```{r}

ggplot(bootstrap_coverage, aes(x = p, y = coverage, color = as.factor(n))) +
  geom_line(size = 1.2) +
  geom_hline(yintercept = 0.95, linetype = "dashed", color = "gray40") +
  labs(
    title = "Bootstrap 95% CI Coverage",
    subtitle = "Empirical coverage by true FT% and sample size",
    x = "True Free Throw Probability (p)",
    y = "Coverage Probability",
    color = "Sample Size (n)"
  ) +
  theme_minimal()
```
```{r}
ggplot(coverage_long, aes(x = p, y = coverage, color = as.factor(n))) +
  geom_line(size = 1) +
  geom_hline(yintercept = 0.95, linetype = "dashed", color = "gray") +
  facet_wrap(~ method, ncol = 1) +
  labs(
    title = "Empirical Coverage of Confidence Intervals",
    x = "True Probability (p)",
    y = "Coverage",
    color = "Sample Size (n)"
  ) +
  theme_minimal()
```
The coverage of the boostrapped interval looks similar to the wald coverage at 10 but more similar to the agresti coull at 100 and 500. We can assume that the coverage is approximately between wald and agresti coull in terms of robustness to n and p.


Lab 11.1.4 

Simulate different alpha and beta values, compare to previously characterized confidence intervals

```{r}
#different values 
a = seq(2, 10, length.out = 20)
b = seq(2, 10, length.out = 20)
beta_grid = expand.grid(a = a, b = b)

posterior_df <- crossing(nba_summary, beta_grid) %>%
  mutate(
    posterior_lower = qbeta(0.025, tot_FTM.x + a, tot_FTA.x - tot_FTM.x + b),
    posterior_upper = qbeta(0.975, tot_FTM.x + a, tot_FTA.x - tot_FTM.x + b),
    width_cred = posterior_upper - posterior_lower
  )

head(posterior_df)
ggplot(posterior_df, aes(x = width_cred, y = FT_percent.x)) +
  geom_point(aes(color = as.factor(a), shape = as.factor(b)), alpha = 0.6) +
  labs(
    title = "Posterior Credible Interval Width vs. Free Throw Percentage",
    x = "Width of Credible Interval",
    y = "Free Throw Percentage",
    color = "Alpha (a)",
    shape = "Beta (b)"
  ) +
  theme_minimal() +
  scale_color_brewer(palette = "PiYg") +
  scale_shape_manual(values = 1:20) +
  theme(legend.position = "none")
```

Compare widths of all the intervals on a plot 

```{r}
ggplot(posterior_df, aes(x = FT_percent.x)) +
  geom_point(aes(y = wald_lower, color = "Wald Lower"), alpha = 0.3) +
  geom_point(aes(y = wald_upper, color = "Wald Upper"), alpha = 0.3) +
  geom_point(aes(y = ac_lower, color = "Agresti-Coull Lower"), alpha = 0.3) +
  geom_point(aes(y = ac_upper, color = "Agresti-Coull Upper"), alpha = 0.3) +
  geom_point(aes(y = boot_lower, color = "Bootstrap Lower"), alpha = 0.3) +
  geom_point(aes(y = boot_upper, color = "Bootstrap Upper"), alpha = 0.3) +
  geom_point(aes(y = posterior_lower, color = "Posterior Lower"), alpha = 0.3) +
  geom_point(aes(y = posterior_upper, color = "Posterior Upper"), alpha = 0.3) +
  labs(
    title = "Comparison of Confidence and Credible Intervals",
    x = "Free Throw Percentage",
    y = "Interval Bounds",
    color = "Interval Type"
  ) +
  theme_minimal() +
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "bottom")
```


