---
title: "09_Maximilian"
author: "Maximilian J. Gebauer"
date: "2025-06-11"
output: html_document
---

```{r Setup, include=FALSE, results='hide', warning=FALSE}
knitr::opts_chunk$set(echo = T, fig.width=8, fig.height=4)
options(scipen = 0, digits = 3)  # controls base R output

# Package setup
if(!require("pacman")) install.packages("pacman")

pacman::p_load(tidyverse, dplyr, ggthemes, data.table, lubridate, glmnet,
               GGally, RColorBrewer, ggsci, plotROC, usmap,
               plotly, ggpubr, vistime, coefplot, skimr, car, ggrepel, slider, lubridate,
               tidymodels,ranger,vip,ggplot2, tune,dials,pdp, purrr, stringr, lmtest,
               sandwich, nnet, doParallel, foreach)
```

```{r}
set.seed(8)
```

```{r}
nba_players <- read.csv("../data/08_nba-free-throws.csv.gz", sep = ";")
```

```{r}
head(nba_players)
```

```{r}
set.seed(8)
agg_data <- nba_players %>%
  group_by(Player) %>%
  mutate(
    Total_FTA = round(sum(FTA * G), digits=0),
    Total_FTM = round(sum(FT * G), digits=0),
    FT_percent = Total_FTM / Total_FTA * 100
  ) %>%
  filter(Total_FTA > 24) %>%
  select(Player, Total_FTA, Total_FTM, FT_percent) %>%
  unique() %>%
  arrange(Total_FTA) %>%
  ungroup()

agg_data$Player <- as.factor(agg_data$Player)

agg_data_sub <- agg_data %>% 
  slice_sample(n=30)


```

```{r, fig.width=10, fig.height=6, warning=FALSE, message=FALSE}
wald_ci <- function(x, n) {
  p <- x / n
  se <- sqrt(p * (1 - p) / n)
  lower <- p - 1.96 * se
  upper <- p + 1.96 * se
  c(lower, upper)
}

# Agresti-Coull 95% CI function
ac_ci <- function(x, n) {
  z <- 1.96
  n_adj <- n + 4
  p_adj <- (x + 2) / n_adj
  se_adj <- sqrt(p_adj * (1 - p_adj) / n_adj)
  lower <- (x/n) - z * se_adj
  upper <- (x/n) + z * se_adj
  c(lower, upper)
}

agg_data_with_ci <- agg_data_sub %>%
  rowwise() %>%
  mutate(
    FT = Total_FTM / Total_FTA,
    wald_low = wald_ci(Total_FTM, Total_FTA)[1] * 100,
    wald_high = wald_ci(Total_FTM, Total_FTA)[2] * 100,
    ac_low = ac_ci(Total_FTM, Total_FTA)[1] * 100,
    ac_high = ac_ci(Total_FTM, Total_FTA)[2] * 100
  ) %>%
  ungroup()

ggplot(agg_data_with_ci, aes(x = reorder(Player, FT_percent), y = FT_percent)) +
  geom_point(color = "black", size = 2) +
  
  # Slight left shift for Wald CI
  geom_errorbar(
    aes(ymin = wald_low, ymax = wald_high),
    width = 0.2, color = "red", alpha = 0.6,
    position = position_nudge(x = -0.2)
  ) +

  # Slight right shift for Agresti-Coull CI
  geom_errorbar(
    aes(ymin = ac_low, ymax = ac_high),
    width = 0.2, color = "blue", alpha = 0.6,
    position = position_nudge(x = 0.2)
  ) +

  coord_flip() +
  labs(
    title = "Free Throw Percentage with 95% Wald and Agresti-Coull Confidence Intervals",
    x = "Player",
    y = "FT%"
  ) +
  theme_minimal()
```

```{r}
p_vec    <- seq(0, 1, length.out = 1000)
p_vec    <- p_vec[p_vec != 0 & p_vec != 1]
n_vec    <- c(10, 50, 100, 250, 500, 1000)
num_sims <- 100

coverage_fn <- function(p, n, num_sims) {
  x <- rbinom(num_sims, size = n, prob = p)
  phat <- x / n
  se_w <- sqrt(phat * (1 - phat) / n)
  wald_l <- phat - qnorm(0.975) * se_w
  wald_u <- phat + qnorm(0.975) * se_w
  cov_w <- mean(wald_l <= p & wald_u >= p)
  
  # Agresti–Coull ("add‐2") CI
  n_tilde <- n + 4
  phat_tilde <- (x + 2) / n_tilde
  se_a <- sqrt(phat_tilde * (1 - phat_tilde) / n_tilde)
  ag_l <- x - qnorm(0.975) * se_a
  ag_u <- x + qnorm(0.975) * se_a
  cov_a <- mean(ag_l <= p & ag_u >= p)
  
  c(coverage_wald = cov_w, coverage_ag = cov_a)
}

# build results data.frame
results <- expand.grid(p = p_vec, n = n_vec) %>%
  rowwise() %>%
  mutate(
    cov = list(coverage_fn(p, n, num_sims))
  ) %>%
  unnest_wider(cov) %>%
  ungroup()

plot_data <- results %>%
  pivot_longer(
    cols      = c(coverage_wald, coverage_ag),
    names_to  = "method",
    values_to = "coverage"
  ) %>%
  # explicitly convert the raw names to exactly the labels you want
  mutate(
    method = case_when(
      method == "coverage_ag" ~ "Wald CI",
      method == "coverage_wald"   ~ "Agresti–Coull CI"
    ),
    n = factor(n)
  )

# coverage vs p, colored by method, facetted by n
ggplot(plot_data, aes(x = p, y = coverage, color = as.factor(method))) +
  geom_line(alpha = 0.8) +
  facet_wrap(~ n, ncol = 3) +
  labs(
    title = "Empirical Coverage of 95% Wald vs Agresti–Coull CIs",
    x = expression(p),
    y = "Coverage Probability",
    color = "CI Method"
  ) +
  geom_hline(yintercept = 0.95, linetype = "dashed") +
  theme_minimal()
```

```{r}
num_red <- 18
num_not_red <- 93
(total <- num_red + num_not_red)
```
```{r}
reds <- rep("red", num_red)
non_reds <- rep("not_red", num_not_red)

all_colors <- c(reds, non_reds)
all_colors <- factor(all_colors, levels = c("red", "not_red"))
```

```{r}
true_p_hat <- num_red / total

z <- 1.96

se_p_hat <- sqrt(true_p_hat * (1 - true_p_hat) / total)

(wald_low <- true_p_hat - z * se_p_hat)
true_p_hat
(wald_high <- true_p_hat + z * se_p_hat)
```

```{r}
true_p_hat <- num_red / total

z <- 1.96

p_hat_star <- (num_red + 2) / (total + 4)

se_p_hat_star <- sqrt(p_hat_star * (1 - p_hat_star) / (total + 4))

(ac_low <- true_p_hat - z * se_p_hat_star)
true_p_hat
(ac_high <- true_p_hat + z * se_p_hat_star)
```

```{r}
set.seed(9)
num_trials <- 1000000

reds <- rep("red", num_red)
non_reds <- rep("not_red", num_not_red)

all_colors <- c(reds, non_reds)
all_colors <- factor(all_colors, levels = c("red", "not_red"))

p_hat_trials <- rep(NA, num_trials)

for (i in 1:num_trials) {
  sample_colors <- sample(all_colors, size = total, replace = TRUE)
  num_reds_sampled <- sum(sample_colors == "red")
  
  p_hat_trials[i] <- num_reds_sampled / total
}
```

```{r,fig.width=10, fig.height=6}}
(lower_boot <- quantile(p_hat_trials, 0.025))
(upper_boot <- quantile(p_hat_trials, 0.975))

hist(p_hat_trials, breaks = 50, main = "Distribution of p-hat from 1,000,000 Trials \n ",
     xlab = "p-hat", col = "lightblue", border = "black")
abline(v = lower_boot, col = "red", lwd = 1)
abline(v = upper_boot, col = "red", lwd = 1)
```


P_Hat: 0.162
Bootstrap Interval: (0.0991, 0.234), Width = 0.135
Wald Interval: (0.0936, 0.231), Width = 0.137
Agresti-Coull Interval: (0.0929, 0.231), Width = 0.138

Widths are nearly identical, which is not unexpected given the observed proportion of reds (0.162) is neither too close to 0 or 1 and the sample size is relatively large (111). I ran the bootstrap simulation with 1,000,000 trials to ensure a stable estimate of the bootstrap confidence interval. For my sample, the realtive coverages of the three are nearly identical.


The below chunk computes bootstrap intervals for each player randomly sampled in the previous Lab.

```{r}
set.seed(8)

num_trials <- 10000

num_players <- nrow(agg_data_sub)
lower_boot <- rep(NA, num_players)
upper_boot <- rep(NA, num_players)

for (i in 1:num_players) {
    player <- agg_data_sub$Player[i]
    total_fta <- agg_data_sub$Total_FTA[i]
    total_ftm <- agg_data_sub$Total_FTM[i]
    
    p_hat <- total_ftm / total_fta
    
    bootstrap_samples <- replicate(num_trials, {
      sample_ftm <- rbinom(1, size = total_fta, prob = p_hat)
      sample_fta <- total_fta
      sample_p_hat <- sample_ftm / sample_fta
      return(sample_p_hat)
    })
    
    lower_boot[i] <- quantile(bootstrap_samples, 0.025)
    upper_boot[i] <- quantile(bootstrap_samples, 0.975)
}

new_interval_df <- agg_data_sub

new_interval_df$lower_boot <- lower_boot
new_interval_df$upper_boot <- upper_boot
```

Below is the plot for the random smaple of 30 players with all three CIs plotted along with their observed proportion of successes. 
```{r, fig.width=10, fig.height=6, warning=FALSE, message=FALSE}
complete_results <- agg_data_with_ci %>%
  left_join(new_interval_df %>% 
              select(Player, lower_boot, upper_boot), 
            by = "Player")

ggplot(complete_results, aes(x = reorder(Player, FT_percent), y = FT_percent)) +
  geom_point(color = "black", size = 2) +
  geom_errorbar(
    aes(ymin = wald_low, ymax = wald_high, color = "Wald"),
    width = 0.2, alpha = 0.6,
    position = position_nudge(x = -0.2),
    show.legend = TRUE
  ) +
  geom_errorbar(
    aes(ymin = ac_low, ymax = ac_high, color = "Agresti–Coull"),
    width = 0.2, alpha = 0.6,
    position = position_nudge(x = 0.2),
    show.legend = TRUE
  ) +
  geom_errorbar(
    aes(ymin = lower_boot * 100, ymax = upper_boot * 100, color = "Bootstrap"),
    width = 0.2, alpha = 0.6,
    position = position_nudge(x = 0.4),
    show.legend = TRUE
  ) +
  scale_color_manual(
    name   = "CI method",
    values = c(
      "Wald"          = "red",
      "Agresti–Coull" = "blue",
      "Bootstrap"     = "green"
    )
  ) +
  coord_flip() +
  labs(
    title = "Free Throw Percentage with 95% Wald, Bootstrap, and Agresti–Coull CIs",
    x     = "Player",
    y     = "FT%"
  ) +
  theme_minimal()
```

Compute all three intervals and their empirical coverage rates in a simulation study

```{r}
p_vec    <- seq(0, 1, length.out = 1000)
p_vec    <- p_vec[p_vec != 0 & p_vec != 1]
n_vec    <- c(10, 50, 100, 250, 500, 1000)
num_sims <- 100
B        <- 500                       
z        <- qnorm(0.975)

coverage_fn <- function(p, n, num_sims, B) {
  x      <- rbinom(num_sims, size = n, prob = p)
  phat   <- x / n
  se_w   <- sqrt(phat * (1 - phat) / n)
  wald_l <- phat - z * se_w
  wald_u <- phat + z * se_w
  cov_w  <- mean(wald_l <= p & wald_u >= p)
  
  n_tilde    <- n + 4
  phat_tilde <- (x + 2) / n_tilde
  phat   <- x / n
  se_a       <- sqrt(phat_tilde * (1 - phat_tilde) / n_tilde)
  ag_l       <- phat - z * se_a
  ag_u       <- phat + z * se_a
  cov_a      <- mean(ag_l <= p & ag_u >= p)
  
  cov_b <- mean(map_lgl(x, function(x_i) {
    data_i    <- c(rep(1, x_i), rep(0, n - x_i))
    phat_star <- replicate(B, mean(sample(data_i, size = n, replace = TRUE)))
    ci        <- quantile(phat_star, probs = c(0.025, 0.975))
    (ci[1] <= p) && (p <= ci[2])
  }))
  
  tibble(
    coverage_wald = cov_w,
    coverage_ag   = cov_a,
    coverage_boot = cov_b
  )
}

grid <- expand.grid(p = p_vec, n = n_vec)

cores <- parallel::detectCores() - 1
cl    <- makeCluster(cores)
registerDoParallel(cl)

results <- foreach(i = seq_len(nrow(grid)), .combine = bind_rows, .packages = c("dplyr","purrr")) %dopar% {
  pi <- grid$p[i]
  ni <- grid$n[i]
  
  coverage_fn(pi, ni, num_sims, B) %>%
    mutate(p = pi, n = ni)
}

stopCluster(cl)

plot_data <- results %>%
  pivot_longer(
    cols      = starts_with("coverage_"),
    names_to  = "method",
    values_to = "coverage"
  ) %>%
  mutate(
    method = case_when(
      method == "coverage_wald" ~ "Wald CI",
      method == "coverage_ag"   ~ "Agresti–Coull CI",
      method == "coverage_boot" ~ "Bootstrap CI"
    ),
    n = factor(n)
  )
```

All three are plotted below. Boot performs similairly to AC with both out performing Wald in cases where n is small and/or p is close to 0 or 1.

```{r}
ggplot(plot_data, aes(x = p, y = coverage, color = method)) +
  geom_line(alpha = 0.8) +
  facet_wrap(~ n, ncol = 3, labeller = label_both) +
  geom_hline(yintercept = 0.95, linetype = "dashed") +
  scale_color_manual(
    name   = "CI Method",
    values = c(
      "Wald CI"          = "red",
      "Agresti–Coull CI" = "blue",
      "Bootstrap CI"     = "yellow"
    )
  ) +
  labs(
    title = "Empirical Coverage of 95% CIs: Wald, Agresti–Coull, and Bootstrap",
    x     = expression(p),
    y     = "Coverage Probability"
  ) +
  theme_minimal()
```






