---
title: "R Notebook"
output: html_notebook
---

12.1.3 Deriving Posteriror Means

```{r}
#Write a formula for posterior mean of normal-normal model 
nba = read.csv('../data/12_nba-box-scores.csv.gz')
head(nba)

library(ggplot2)
library(tidyverse)

#define set 
nba_set <- nba %>%
  group_by(namePlayer) %>%
  filter(n() > 30) %>%    # use n() instead of length()
  ungroup()

nba_average <- nba_set %>%
  group_by(namePlayer) %>%
  mutate(
    meanPoints = mean(pts, na.rm = TRUE),
    varPoints = var(pts, na.rm = TRUE),
    g = n()
  ) %>%
  ungroup()


```

Task 4 - estimate sigma$^2$
```{r}
nba_variance <- nba_average %>% 
  group_by(namePlayer) %>% 
  mutate(
    variance = var(pts/possessions)
  ) %>% 
  ungroup()

mu_hat <- nba_variance %>%
  summarize(mean_points = sum(pts) / sum(possessions)) %>%
  pull(mean_points)

average_players = nba_variance %>% 
  filter(8 < meanPoints & meanPoints < 12)

sigma2_hat = average_players %>% 
  summarize(sigma2 = mean(variance, na.rm = TRUE)) %>% 
  pull(sigma2)

print(sigma2_hat)

```

Task 5: Estimate T$^2$

```{r}
tau2_grid <- 10^seq(-7, -3, length.out = 10)

# Split each player’s career into two halves
split_players <- nba_variance %>% 
  group_by(namePlayer) %>% 
  arrange(dateGame) %>% 
  mutate(
    idx = row_number(),
    G_i = n(),
    half = floor(G_i / 2),
    which = if_else(idx <= half, "first_half", "second_half")
  ) %>% 
  summarize(
    m_ik = mean(pts[which == "first_half"] / possessions[which == "first_half"], na.rm = TRUE),
    d2_i = list(pts[which == "second_half"] / possessions[which == "second_half"]),
    k_i = first(half),
    .groups = "drop"
  )

# Compute RMSE for each tau²
rmse_results <- sapply(tau2_grid, function(tau2) {
  per_player_rmse <- mapply(function(m_ik, d2_i, k_i) {
    mu_ik <- (sigma2_hat / (sigma2_hat + k_i * tau2)) * mu_hat +
             (k_i * tau2 / (sigma2_hat + k_i * tau2)) * m_ik
    sqrt(mean((mu_ik - unlist(d2_i))^2, na.rm = TRUE))
  },
  m_ik = split_players$m_ik,
  d2_i = split_players$d2_i,
  k_i  = split_players$k_i,
  SIMPLIFY = TRUE)
  
  mean(per_player_rmse, na.rm = TRUE)
})

# Store results
tau2_results <- tibble(
  tau2 = tau2_grid,
  mean_RMSE = rmse_results
)

# Identify best tau²
best_tau2 <- tau2_results %>%
  filter(mean_RMSE == min(mean_RMSE))

best_tau2

```
```{r}
#Visualize 
ggplot(tau2_results, aes(x = tau2, y = mean_RMSE)) +
  geom_line() +
  geom_point() +
  scale_x_log10() +
  labs(title = "Empirical Bayes RMSE vs τ²",
       x = expression(tau^2),
       y = "Mean RMSE") +
  theme_minimal()
```
Compute posterior means
```{r}

nba_variance_posterior <- nba_variance %>% 
  group_by(namePlayer) %>%
  arrange(dateGame) %>% 
  mutate(
    X = pts / possessions,
    sigma2_hat = sigma2_hat,
    tau2_hat = best_tau2$tau2
  ) %>% 
  ungroup()

nu2_hat <- nba_variance_posterior %>%
  group_by(namePlayer) %>%
  summarize(mean_X = mean(X, na.rm = TRUE)) %>%
  summarize(nu2_hat = var(mean_X, na.rm = TRUE)) %>%
  pull(nu2_hat)

print(nu2_hat)

mean(nba_variance_posterior$X, na.rm = TRUE)
  
```

Plot 
```{r}

set.seed(12345)
nba_posterior <- nba_variance_posterior %>%
  arrange(namePlayer, dateGame) %>% 
  group_by(namePlayer) %>% 
  mutate(
    mu_bij = NA_real_,
    game_num = row_number()
  ) %>% 
  group_split(.keep=TRUE)
  
for (i in seq_along(nba_posterior)) {
  player_df <- nba_posterior[[i]]
  n_games <- nrow(player_df)
  X_j <- player_df$pts[j] / player_df$possessions[j]
  mu_vec <- numeric(n_games)

  # First game: use global prior
  mu_vec[1] <- ((player_df$pts[1] / (sigma2_hat * player_df$possessions[1])) + (mu_hat / nu2_hat)) /
               ((1 / (sigma2_hat * player_df$possessions[1])) + (1 / nu2_hat))
  
  # Subsequent games: update using previous mu_vec[j - 1]
  if (n_games > 1) {
    for (j in 2:n_games) {
      mu_vec[j] <-mu_vec[j] <- (X_j / sigma2_hat + mu_vec[j - 1] / .001) /
             (1 / sigma2_hat + 1 / .001)
    }
  }

  nba_posterior[[i]]$mu_bij <- mu_vec
}

nba_posterior_all <- bind_rows(nba_posterior )

sample_players <- nba_posterior_all %>%
  distinct(namePlayer) %>%
  slice_sample(n = 10) %>%
  pull(namePlayer)

nba_sample_df <- nba_posterior_all %>%
  filter(namePlayer %in% sample_players)

# Plot
library(ggplot2)
nba_sample_df %>%
  arrange(namePlayer, dateGame) %>%
  group_by(namePlayer) %>%
  mutate(game_number = row_number()) %>%
  ggplot(aes(x = game_number, y = mu_bij, color = namePlayer)) +
  geom_line(size = 1.2) +
  labs(
    title = "Posterior Scoring Quality Over Time",
    x = "Game Number in Career",
    y = expression(mu[bij])
  ) +
  theme_minimal() 

```

 
12.2.1 
Task 1 - fit field goal probability as function of ydl
```{r}
library(splines)
nfl = read.csv('../data/12_field-goals.csv.gz')
head(nfl)

model = glm(fg_made ~ ydl, data = nfl, family = binomial(link = "logit"))
model_spline <- glm(fg_made ~ bs(ydl, df = 4),  # df = degrees of freedom
                    data = nfl,
                    family = binomial(link = "logit"))
summary(model)
summary(model_spline)
```
```{r}
#set alpha = 0.95

#use model to predict
nfl <- nfl %>%
  mutate(
    p0 = predict(model_spline, newdata = nfl, type = "response"),
    FGPA = fg_made - p0
  )

compute_kicker_quality <- function(data, alpha = 0.9) {
  data <- data %>%
    arrange(kicker) %>%
    group_by(kicker) %>%
    mutate(kick_num = row_number()) %>%
    ungroup()

  # Preallocate column for kicker quality
  data$KQ <- 0

  # Loop over each kicker
  kicker_list <- unique(data$kicker)

  for (k in kicker_list) {
    kicker_data <- data[data$kicker == k, ]
    n_kicks <- nrow(kicker_data)
    
    if (n_kicks < 2) next

    kq_vec <- numeric(n_kicks)
    kq_vec[1] <- 0  # KQ_i1 = 0

    for (j in 2:n_kicks) {
      kq_vec[j] <- alpha * kq_vec[j - 1] + kicker_data$FGPA[j - 1]
    }

    data$KQ[data$kicker == k] <- kq_vec
  }

  return(data)
}

nfl_kq <- compute_kicker_quality(nfl, alpha = 0.95)

# Sample a few kickers to visualize
sample_kickers <- nfl_kq %>%
  distinct(kicker) %>%
  slice_sample(n = 10) %>%
  pull(kicker)

# Plot trajectories
nfl_kq %>%
  filter(kicker %in% sample_kickers) %>%
  ggplot(aes(x = kick_num, y = KQ, color = kicker)) +  # color per kicker
  geom_line(size = 1.2) +
  labs(
    title = "Kicker Quality Over Career (α = 0.95)",
    x = "Kick Number",
    y = "Kicker Quality (KQ)",
    color = "Kicker"
  ) +
  theme_minimal()
```

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Cmd+Shift+K* to preview the HTML file). 

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.

