---
title: "13 - Shrinkage Estimation"
output: html_notebook
---

Task 1 - fit equations 

```{r}
putting = read.csv('../data/13_putts-train.csv.gz')
head(putting)

#equation 13.1 - global mean 
p1 = mean(putting$X)
print(p1) 

putting <- putting %>% 
  mutate(
    p1 = p1
  )

head(putting)
```
We see that the average putting% for all golfers is 0.403766 

Equation 2, the MLE, is just the column of X already in the dataset 

```{r}
#equation 3 

#define constant C for all players
putting <- putting %>% 
  mutate(
    C = mean(X)*(1-mean(X)),
    Xi_tilde = X/(sqrt(C/N)), 
    Xi_tilde_bar = mean(Xi_tilde),
    Sx2 = sum((Xi_tilde - Xi_tilde_bar)^2),
    theta_em =Xi_tilde_bar + (1-((length(putting)-1)/Sx2)) * (Xi_tilde - Xi_tilde_bar),
    p3 = sqrt(C/N) * theta_em
  )

head(putting)
```

We see that the estimates are similar to the MLE, with pretty much the same numbers for the first three digits

```{r}
#equation 4 - variance stabilizing transformation

k = length(putting) 

putting <- putting %>% 
  mutate(
    H = X * N, #add column for hits per each golfer
    Xi_doubletilde = asin(sqrt((H+3/8)/(N+3/4))), 
    nu_i2 = 1/(4*N),
    nu = sqrt(nu_i2),
    new_Xi_tilde = Xi_doubletilde/nu,
    new_Xi_tilde_bar = mean(new_Xi_tilde),
    new_Sx2 = sum((new_Xi_tilde - new_Xi_tilde_bar)^2),
    theta_em2 = new_Xi_tilde_bar +  (1-(k-1)/new_Sx2) *(new_Xi_tilde - new_Xi_tilde_bar),
    p4 = (sin(nu*theta_em2))^2
  )


putting_cleaned <- putting %>% 
  select(Player, N, X, p1, p3, p4)

head(putting_cleaned)
```

We see that values are still relatively similar to p2, the MLE, but the digits are slightly more divergent after the thousandths

```{r}
#equation 5 
putting <- putting %>% 
  mutate(
    Sx2_5 = sum((X - mean(X))^2) / (k - 1),
    tau2 = var(X) - C*mean(1/N),
    p5 = p1 + tau2/(tau2+(C/N))*(X-p1)
  )

putting_cleaned <- putting %>% 
  select(Player, N, X, p1, p3, p4, p5)
head(putting)
head(putting_cleaned)
```
```{r}
#equation 6 
putting <- putting %>% 
  mutate(
    Xi_tilde_6 =asin(sqrt((H+ 3/8) / (N + 3/4))),
    Xi_tilde_6_bar = mean(Xi_tilde_6),
    Sx2_6 = sum((Xi_tilde_6 - Xi_tilde_6_bar)^2)/(k-1),
    new_tau2 = Sx2_6 - mean(nu_i2),
    theta_em2_hat = Xi_tilde_6_bar + (new_tau2/(new_tau2+nu_i2))*(Xi_tilde_6 - Xi_tilde_6_bar),
    p6 = (sin(theta_em2_hat))^2
  )

putting_cleaned <- putting %>% 
  select(Player, N, X, p1, p3, p4, p5, p6)

head(putting_cleaned)
```
Task 2 - compare estimates to test data with MSE 

```{r}
putting_test = read.csv('../data/13_putts-test.csv.gz')
mse <- function(predicted, actual) {
  mean((predicted - actual)^2)
}

#compute MSE for each 

putting_results <- bind_cols(putting_cleaned, X_test = putting_test$X)

tibble(
  model = c("p1", "p2", "p3", "p4", "p5", "p6"),
  mse = c(
    mse(putting_results$p1, putting_results$X_test),
    mse(putting_results$X,   putting_results$X_test),
    mse(putting_results$p3,  putting_results$X_test),
    mse(putting_results$p4,  putting_results$X_test),
    mse(putting_results$p5,  putting_results$X_test),
    mse(putting_results$p6,  putting_results$X_test)
  )
)

```
Task 3 - Erm for some reason the global mean is the best, then mse4, then mse6, then mse5, mse3, and then the MLE

Task 4: Plot predicted performances against each estimate
```{r}
putting_cleaned <- putting_cleaned %>% 
  mutate(result = putting_test$X)
  
putting_long <- putting_cleaned %>%
  select(-N, -Player) %>%  # drop column N
  pivot_longer(cols = starts_with("p"), names_to = "prediction_set", values_to = "predicted")

# Now plot each prediction line against the test value X
ggplot(putting_long, aes(x = result, y = predicted, color = prediction_set)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, color = "black", linetype = "dashed") +
  labs(title = "Model Predictions vs True X",
       x = "True Test Value (X)",
       y = "Predicted Value",
       color = "Prediction Set") +
  facet_wrap(~prediction_set) +
  theme_minimal()
```

```{r}

```

