---
title: "14 - fully bayesian models"
output: html_notebook
---

Preliminary Data Cleaning

```{r}

library(rstan)
library(tidyverse)
library(ggplot2)

nfl_data = read.csv("../data/14_nfl-games.csv.gz")
head(nfl_data)
teams <- sort(unique(c(nfl_data$home_team, nfl_data$away_team)))
team_indices <- setNames(seq_along(teams), teams)

# 3. Assign numeric team indices
nfl_data <- nfl_data %>%
  mutate(
    H = team_indices[home_team],
    A = team_indices[away_team]
  )

nfl_data <- nfl_data %>% 
  mutate(s_i = season - 2017,
         home_team = as.factor(home_team),
         away_team = as.factor(away_team),
         home_win = (total_home_score - total_away_score)
           )

head(nfl_data)
```

Task 1 - fit model

```{r}
stan_model <- stan_model(file = "14_victoria.stan")
data_train <- list(
  N_games = nrow(nfl_data),
  N_teams = length(team_indices),
  N_seasons = length(unique(nfl_data$season)),
  y = nfl_data$pts_H_minus_A,
  H = nfl_data$H,
  A = nfl_data$A,
  S = nfl_data$s_i
)
# train the model
fit = sampling(stan_model, data = data_train , iter = 1500, chains = 1, seed = 12345)
fit
```

```{r}
posterior <- as.matrix(fit)
alpha_samples <- as.matrix(fit, pars = "alpha")
beta_colnames <- grep("beta\\[[0-9]+,6\\]", colnames(posterior), value = TRUE)

# Subset just the columns for the latest season


beta_samples <- posterior[, beta_colnames]

# Check dimensions
str(beta_samples)  # Should be [iterations x teams]

# Summary table
beta_summary <- tibble(
  team = names(team_indices),  # should match Stan team order
  mean = apply(beta_samples, 2, mean),
  lower = apply(beta_samples, 2, quantile, 0.025),
  upper = apply(beta_samples, 2, quantile, 0.975)
)

# Plot team strength
ggplot(beta_summary, aes(x = reorder(team, mean), y = mean)) +
  geom_point() +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.2) +
  coord_flip() +
  labs(
    title = "Posterior Means of Team Strength (Most Recent Season)",
    x = "Team",
    y = "Estimated Team Strength (β)"
  ) +
  theme_minimal()

# Check structure

```
```{r}
str(alpha_samples)

alpha_summary <- tibble(
  team = names(team_indices),
  mean = apply(alpha_samples, 2, mean),
  lower = apply(alpha_samples, 2, quantile, 0.025),
  upper = apply(alpha_samples, 2, quantile, 0.975)
)

ggplot(alpha_summary, aes(x = reorder(team, mean), y = mean)) +
  geom_point() +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.2) +
  coord_flip() +
  labs(
    title = "Posterior Means of Team-Specific Home Field Advantage",
    x = "Team",
    y = "Estimated Home Field Advantage (α_j)"
  ) +
  theme_minimal()
```

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Cmd+Shift+K* to preview the HTML file). 

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.

