---
title: "Lab 16 - Bias and Variance Tradeoff"
output: html_notebook
---

Task 1 - true parameter vector beta according to following distributions

```{r}
#load in data
library(tidyverse)
library(ggplot2)
park_effects <- read_csv("../data/16_park-effects.csv.gz")
```

```{r}
beta0 = .4

n_parks = length(unique(park_effects$PARK))
n_off = length(unique(park_effects$OT_YR))
n_def = length(unique(park_effects$DT_YR))

beta_parks = rnorm(n_parks, mean = 0.04, sd = .065)
beta_off = rnorm(n_off, mean = 0.02, sd = 0.045)
beta_def = rnorm(n_def, mean = 0.03, sd = 0.07)

beta = c(beta0, beta_parks, beta_off, beta_def)
```

2 - encode X matrix
```{r}
park_effects <- park_effects %>%
  mutate(row_id = row_number())

# One-hot encode PARK
park_wide <- park_effects %>%
  select(row_id, PARK) %>%
  mutate(value = 1) %>%
  pivot_wider(
    names_from = PARK,
    values_from = value,
    values_fill = list(value = 0),
    names_prefix = "PARK_"
  )

ot_wide <- park_effects %>%
  select(row_id, OT_YR) %>%
  mutate(value = 1) %>%
  pivot_wider(
    names_from = OT_YR,
    values_from = value,
    values_fill = list(value = 0),
    names_prefix = "OT_"
  )

# One-hot encode DT_YR
dt_wide <- park_effects %>%
  select(row_id, DT_YR) %>%
  mutate(value = 1) %>%
  pivot_wider(
    names_from = DT_YR,
    values_from = value,
    values_fill = list(value = 0),
    names_prefix = "DT_"
  ) 



# Combine into full matrix X
X <- park_wide %>%
  left_join(ot_wide, by = "row_id") %>%
  left_join(dt_wide, by = "row_id") %>%
  mutate(Intercept = 1) %>%
  select(row_id, Intercept, everything(), -row_id)
  
```
 
3 - fit to betas
```{r}
library(truncnorm)
n_sims = 100

y_list <- vector("list", n_sims)

beta <- as.numeric(beta)
beta <- matrix(beta, ncol = 1)
X <- as.matrix(X)

for (m in 1:n_sims) {
  mu <- X %*% beta  # linear predictor: n x 1 vector
  y_m <- rtruncnorm(n, a = 0, mean = mu, sd = 1)  # Truncated Normal
  y_list[[m]] <- round(y_m)  # Round to integer
}
```

4 - fit OLS, Ridge, and LASSO regression
```{r}
#prepare empty matrices for all models 
library(glmnet)

beta_ols_mat <- matrix(NA, nrow = n_sims, ncol = n_parks)
beta_ridge_mat <- matrix(NA, nrow = n_sims, ncol = n_parks)
beta_lasso_mat <- matrix(NA, nrow = n_sims, ncol = n_parks)
lambdas <- 10^seq(-3, 3, by = 0.2)

#for loop to fit models
for (m in 1:n_sims) {
  y_m <- y_list[[m]]

  # Fit OLS
  model_ols <- lm(y_m ~ ., data = as.data.frame(X))
  beta_ols_mat[m, ] <- coef(model_ols)[2:(n_parks + 1)]  # skip intercept

  # Fit Ridge
  model_ridge <- cv.glmnet(X, y_m, alpha = 0, lambda = lambdas)
  ridge_coefs <- as.vector(coef(model_ridge, s = "lambda.min"))[-1]  # drop intercept
  beta_ridge_mat[m, ] <- ridge_coefs[1:n_parks]

  # Fit LASSO
  model_lasso <- cv.glmnet(X, y_m, alpha = 1, lambda = lambdas)
  lasso_coefs <- as.vector(coef(model_lasso, s = "lambda.min"))[-1]  # drop intercept
  beta_lasso_mat[m, ] <- lasso_coefs[1:n_parks]
}

beta_ols_mat <- beta_ols_mat[1:50, 2:30] #adjustment since first column is NA

```
 
5 and 6 - get the bias and variance of the estimates

```{r}

beta_ols_mat <- beta_ols_mat[1:50, ]
beta_ridge_mat <- beta_ridge_mat[1:50, ]
beta_lasso_mat <- beta_lasso_mat[1:50, ]


beta_ols_mean <- colMeans(beta_ols_mat)
beta_ridge_mean <- colMeans(beta_ridge_mat)
beta_lasso_mean <- colMeans(beta_lasso_mat)

# Bias
bias_ols <- sqrt(sum((beta_parks - beta_ols_mean)^2))
bias_ridge <- sqrt(sum((beta_parks - beta_ridge_mean)^2))
bias_lasso <- sqrt(sum((beta_parks - beta_lasso_mean)^2))

bias = c(bias_ols, bias_ridge, bias_lasso)

# Variance
var_ols <- mean(apply(beta_ols_mat, 2, var))
var_ridge <- mean(apply(beta_ridge_mat, 2, var))
var_lasso <- mean(apply(beta_lasso_mat, 2, var))

variance = c(var_ols, var_ridge, var_lasso)

cat("Bias (OLS):", round(bias_ols, 4), "\n")
cat("Bias (Ridge):", round(bias_ridge, 4), "\n")
cat("Bias (LASSO):", round(bias_lasso, 4), "\n")
cat("Variance (OLS):", round(var_ols, 4), "\n")
cat("Variance (Ridge):", round(var_ridge, 4), "\n")
cat("Variance (LASSO):", round(var_lasso, 4), "\n")

bias_var_df <- tibble(
  model = c("OLS", "Ridge", "LASSO"),
  bias = bias,
  variance = variance
)

ggplot(bias_var_df, aes(x = model, y = bias, fill = model)) +
  geom_col() +
  labs(
    title = "Average Bias of Park Effect Estimates",
    y = "Bias",
    x = "Model"
  ) +
  scale_fill_manual(values = c("OLS" = "lightblue4", "Ridge" = "lightcoral", "LASSO" = "palegoldenrod")) +
  theme_minimal()

# Variance plot
ggplot(bias_var_df, aes(x = model, y = variance, fill = model)) +
  geom_col() +
  labs(
    title = "Average Variance of Park Effect Estimates",
    y = "Variance",
    x = "Model"
  ) +
  scale_fill_manual(values = c("OLS" = "lightblue4", "Ridge" = "lightcoral", "LASSO" = "palegoldenrod")) +
  theme_minimal()

```

6 - Visualization
```{r}
boxplot(as.data.frame(beta_ols_mat), main = "OLS Park Effect Estimates", col = 'lightblue4')
boxplot(as.data.frame(beta_ridge_mat), main = "Ridge Park Effect Estimates", col = 'lightcoral')
boxplot(as.data.frame(beta_lasso_mat), main = "LASSO Park Effect Estimates", col = 'palegoldenrod')
```
We see that the OLS estimates are much more uniform across parks. Ridge and LASSO both show a good amount of variability however the LASSO estimates are generally more centered but with longer tails 

We can also use density plots to visualize
```{r}

beta_df <- bind_rows(
  as.data.frame(beta_ols_mat) %>% pivot_longer(everything(), names_to = "park", values_to = "coef") %>% mutate(model = "OLS"),
  as.data.frame(beta_ridge_mat) %>% pivot_longer(everything(), names_to = "park", values_to = "coef") %>% mutate(model = "Ridge"),
  as.data.frame(beta_lasso_mat) %>% pivot_longer(everything(), names_to = "park", values_to = "coef") %>% mutate(model = "LASSO")
)

ggplot(beta_df, aes(x = coef, fill = model)) +
  geom_density(alpha = 0.4) +
  facet_wrap(~ park, scales = "free") +
  labs(title = "Density of Park Effect Estimates", x = "Estimated Coefficient", y = "Density") +
  theme_minimal()+
  scale_fill_manual(values = c("OLS" = "lightblue4", "Ridge" = "lightcoral", "LASSO" = 'palegoldenrod')) 
```

We can also look at violin plots of the coeffs 

```{r}
ggplot(beta_df, aes(x = model, y = coef, fill = model)) +
  geom_violin(alpha = 0.5) +
  facet_wrap(~ park, scales = "free_y") +
  labs(title = "Violin Plots of Park Coefficients by Model") +
  theme_minimal() +
  scale_fill_manual(values = c("OLS" = "lightblue4", "Ridge" = "lightcoral", "LASSO" = 'palegoldenrod')) 

```
We can see that the OLS estimates have much greater variance than the other two models 
