---
title: "19_adam"
output:
  pdf_document: default
  html_document: default
date: "2025-06-30"
---

```{r setup, include=FALSE}
#############
### SETUP ###
#############

library(cluster)      # for hierarchical clustering and silhouette analysis
library(factoextra)   # for clustering visualization and elbow method
library(flexclust)    # for k-means clustering
library(ggdendro)     # for dendrogram visualization
library(ggplot2)      # for plotting
library(mclust)       # for Gaussian Mixture Models
library(tidyverse)    # for data manipulation and visualization
library(dplyr)         
library(tidyr)

# set seed
set.seed(19)

####################
### SPOTIFY DATA ###
####################

# read in data
spotify_data = read_csv("../data/19_spotify-train.csv.gz")

# preview data
head(spotify_data)
```
here we have a lovely lil bar plot of all genres>3 count
```{r, echo = FALSE}
# Step 1: Separate multiple genres into individual rows
genre_counts <- spotify_data %>%
  separate_rows(Genre, sep = ";") %>%       # splits on ";"
  mutate(Genre = str_trim(Genre)) %>%       # removes leading/trailing spaces
  count(Genre, sort = TRUE) %>% 
  filter(n>3)

# Step 2: Plot the genre counts


ggplot(genre_counts, aes(x = reorder(Genre, -n), y = n)) +
  geom_bar(stat = "identity", fill = "darkgreen") +
  labs(title = "Number of Songs by Genre (n>3)",
       x = "Genre", y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
lets ermmmm do a lil correlation heatmap. also standardized data in here

```{r, echo = FALSE}
library(ggcorrplot)

# Drop non-numeric columns (like Song Name, Artist, etc.)
spotify_numeric <- spotify_data %>%
  select(where(is.numeric))

# (Optional) Standardize the data
spotify_scaled <- scale(spotify_numeric)

cor_matrix <- cor(spotify_scaled)

ggcorrplot(cor_matrix, 
           type = "lower", 
           lab = TRUE, 
           title = "Correlation Heatmap of Song Features")


#gotta add back artist and track
# Convert scaled matrix back to a data frame
scaled_df <- as.data.frame(spotify_scaled)

# Add back the track and artist columns
scaled_df$track <- spotify_data$track
scaled_df$`Artist Name(s)` <- spotify_data$`Artist Name(s)`

# Optional: Reorder so track and artist come first
scaled_df <- scaled_df %>%
  relocate(track, `Artist Name(s)`)


```

clus-turd

```{r}
# Maximum number of clusters to try
max_k <- 10

# Vector to store total within-cluster sum of squares for each k
wcss <- numeric(max_k)


for (k in 1:max_k) {
  km <- kmeans(spotify_numeric, centers = k, nstart = 25)  # Forgy + 10 random starts
  wcss[k] <- km$tot.withinss
}

# Plot the Elbow graph
plot(1:max_k, wcss, type = "b",
     xlab = "Number of clusters (k)",
     ylab = "Total Within-Cluster Sum of Squares (WCSS)",
     main = "Elbow Method for Determining Optimal k")


```

ima make k = 3 since 2 seems super small and not interesting though that is where the elbow is

```{r}
k = 3


clusters = kmeans(spotify_numeric, centers = k, nstart = 25)

scaled_df$cluster <- factor(clusters$cluster)
# Plot the clusters
ggplot(scaled_df, aes(x = `Danceability`, y = `Energy`, color = cluster)) +
  geom_point(alpha = 0.6) +
  labs(title = "K-Means Clustering of Spotify Songs",
       x = "Danceability", y = "Energy") +
  theme_minimal() +
  scale_color_brewer(palette = "Set1")

```
im boutta pop off with this graph just watch

```{r}
numeric_columns <- colnames(scaled_df)[sapply(scaled_df, is.numeric)]

# Calculate means by cluster for all numeric variables
cluster_means <- scaled_df %>%
  group_by(cluster) %>%
  summarise(across(all_of(numeric_columns), mean))

# Reshape to long format for plotting
cluster_means_long <- cluster_means %>%
  pivot_longer(cols = -cluster, names_to = "Variable", values_to = "Average")

ggplot(cluster_means_long, aes(x = cluster, y = Average, fill = cluster)) +
  geom_col(position = position_dodge()) +
  facet_wrap(~ Variable, scales = "free_y") +
  labs(title = "Average Values by Cluster for All Variables",
       x = "Cluster",
       y = "Average Value",
       fill = "Cluster") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_brewer(palette = "Set2")



```

ok so 3 is spivey?? slower, softer, less energy, longer, not rly danceable, more acoustic, more instrumental

1 is more upbeat faster songs, more electronic ones, but less popular than 2. 2 is slower and pretty middle on most categories except popularity. 2 is prob aiwen a lot. eh shes also probably in 1 a bunch but more 2. one i think has the rap/hiphop

```{r}








