---
title: "Max Spotify Predictive Model"
author: "Maximilian J. Gebauer"
date: "2025-06-30"
output: html_document
---


Guide: load packages chunk, (make sure train data file path is correct), then run the following chunk which fits all three of the component models on the training data. In the chunk after, put in the csv file path in "test_data <- read.csv()" object already in code chunk, then run the chunk to pre-process the data for the form my models expect. Then run final chunk which returns a 10 by j matrix where each entry is the jth songs predicted prob for the ith person (and log loss of my model on the test data).


```{r Setup, include=FALSE, results='hide', warning=FALSE}
knitr::opts_chunk$set(echo = T, fig.width=8, fig.height=4)
options(scipen = 0, digits = 3) 

if(!require("pacman")) install.packages("pacman")

pacman::p_load(tidyverse, dplyr, ggthemes, data.table, lubridate, glmnet,
               GGally, RColorBrewer, ggsci, plotROC, usmap,
               plotly, ggpubr, vistime, coefplot, skimr, car, ggrepel, slider, lubridate,
               tidymodels,ranger,vip,ggplot2, tune,dials,pdp, purrr, stringr, lmtest,
               sandwich, xgboost, future.apply, mclust, ggdendro, flexclust, factoextra, cluster,
               parsnip)
```

```{r}
#load training data
data <- read.csv("../data/19_spotify-train.csv.gz")
```

```{r}
#one-hot encode all unique genre values
data_onehot <- data %>%
  mutate(.row = row_number()) %>%
  separate_rows(Genre, sep = ";") %>%
  mutate(Genre = str_trim(Genre)) %>%
  mutate(present = 1) %>%
  pivot_wider(
    id_cols        = .row,
    names_from     = Genre,
    values_from    = present,
    values_fill    = list(present = 0)
  ) %>%
  select(-.row)

#remove unnecessary features
sub_data <- data %>% 
  select(-c(Track.URI,track,Album.Name,Artist.Name.s.,Release.Date,Added.At,Record.Label,Genre, Key))

#combine one hot features and other features
full_data <- cbind(sub_data,data_onehot)

row.names(full_data) <- NULL

#set variable natures
full_data$Explicit <- factor(full_data$Explicit)
full_data$Mode <- factor(full_data$Mode)
full_data$Time.Signature <- factor(full_data$Time.Signature)
full_data$Added.by <- factor(full_data$Added.by)

#features to scale
scaling_features <- c("Duration..ms.", "Popularity", "Danceability", "Energy", "Loudness", "Speechiness", "Acousticness", "Instrumentalness", "Liveness", "Valence", "Tempo")

#pull scaling features
numerics <- full_data[, scaling_features]

#scale requisite features
numerics_scaled <- as.data.frame(scale(numerics))

#pull unscaled features
full_sub_data <- full_data %>%
  select(-scaling_features)

#put categroical, scaled, and one-hot features
cat_data <- cbind(full_sub_data,numerics_scaled)
row.names(cat_data) <- NULL

#set up model matrix to tune multinomial ridge model
set.seed(17761)
X <- model.matrix(Added.by ~ ., data = cat_data)[,-1]
y <- cat_data$Added.by

#set for parallel processing
doParallel::registerDoParallel(cores = 6)

#run grid search over lambda 
fit.lasso <- glmnet::cv.glmnet(X,y,alpha=0,nfolds = 6,family="multinomial", parallel = TRUE)

#stop parallel
doParallel::stopImplicitCluster()

#pull conservative 1SE lambda that minimizes average multinomial deviance across the kfolds
best_s_1 <- fit.lasso$lambda.1se

#pull coefs to check (for fun)
coef_list <- coef(fit.lasso, s = best_s_1)

#refit on optimal lambda
preds_prob <- predict(
  fit.lasso, 
  newx = X, 
  s    = best_s_1, 
  type = "response"
)

#below lines get probs into a df form for later use in an ensamble "stack" 
prob_mat <- preds_prob[,,1] 

prob_mat <- preds_prob[,,1]  

prob_df_1 <- as.data.frame(preds_prob)

k <- 2                    
names(prob_df_1) <- substr(names(prob_df_1), 1, nchar(names(prob_df_1)) - k)


#now setting up random forrest hyperparameter tuning
library(tidymodels)
library(ranger)
library(future)

#re specify data to be safe
X <- model.matrix(Added.by ~ ., cat_data)[ , -1]
y <- cat_data$Added.by

#set data in a matrix ready for training
dtrain <- data.frame(
  Matrix::as.matrix(X),
  Added.by = factor(y)
)

#specify functional form
rf_rec <- recipe(Added.by ~ ., data = dtrain)

#check raw n features
n_feat <- ncol(X)

#tuning number of trees, min leaf size, and and num features to consider per split
rf_spec <- rand_forest(
  trees = tune(),         
  mtry  = tune(),         
  min_n = tune()         
) %>% 
  set_mode("classification") %>% 
  set_engine(
    "ranger",
    importance      = "impurity",
    sample.fraction = tune(),           
    num.threads     = parallel::detectCores()
  )

#specify the tidy models workflow
rf_wf <- workflow() %>% 
  add_recipe(rf_rec) %>% 
  add_model(rf_spec)

#setting conservative ranges for hyperparameters
rf_param <- extract_parameter_set_dials(rf_wf) %>% 
  update(
    trees = trees(c(500, 2000)),
    mtry  = mtry(c(
               max(2L, round(0.05 * n_feat)),   
               max(2L, round(0.30 * n_feat))    
             )),
    min_n = min_n(c(10L, 50L)),
    `sample.fraction` = sample_prop(c(0.50, 0.80))  
  )

#runnign grid search over hyperparameters value
set.seed(17761)
rf_grid <- grid_latin_hypercube(rf_param, size = 60)    

#cross validate for tuning
cv_folds <- vfold_cv(dtrain, v = 10, strata = Added.by)

plan(multisession, workers = parallel::detectCores())

#rune tune
rf_res <- tune_grid(
  rf_wf,
  resamples = cv_folds,
  grid      = rf_grid,
  metrics   = metric_set(mn_log_loss),  
  control   = control_grid(save_pred = TRUE)
)

plan(sequential)

#pull best model based on log loss
best_rf   <- select_best(rf_res)
final_rf  <- finalize_workflow(rf_wf, best_rf) %>% fit(dtrain)

#predict on train, then make sure df is set correctly for stacking
preds2 <- predict(final_rf, dtrain, type="prob")

k <- 6                       
colnames(preds2) <- substring(names(preds2), k + 1)

prob_df_2 <- preds2

#specify response for stacking model
y <- cat_data$Added.by

#ensure col ordering for both sub model dfs is correct
colnames(prob_df_2)  <- levels(y)
colnames(prob_df_1) <- levels(y)

#combine both predicted prob results into one df
stack_df <- data.frame(
  prob_df_2,
  prob_df_1    
)

#set names for specificty, assumes 10 names
names(stack_df) <- paste0(rep(c("rf_", "glm_"), each = 10),
                          rep(head(levels(y), 10), 2))
#fit final stacking model trained on sub model predictions
set.seed(17761)
stack_fit <- cv.glmnet(
  x            = as.matrix(stack_df),
  y            = y,
  family       = "multinomial",
  alpha        = 0,
  type.measure = "deviance",
  nfolds       = 10
)

#pull conservative 1SE lambda for stack later on test data
best_s <- stack_fit$lambda.1se
```


 
```{r}
# test data processing chunk #

#Feed me test data       #########################
test_data <- read.csv("../data/19_spotify-test.csv.gz")  ## <- Put in test data ##
                         #########################

#one hot encode all values found in the Genre feature
test_onehot <- test_data %>%
  mutate(.row = row_number()) %>%
  separate_rows(Genre, sep = ";") %>%
  mutate(Genre = str_trim(Genre)) %>%
  mutate(present = 1) %>%
  pivot_wider(
    id_cols        = .row,
    names_from     = Genre,
    values_from    = present,
    values_fill    = list(present = 0)
  ) %>%
  select(-.row)

# take out unnecessary features
sub_test <- test_data %>% 
  select(-c(Track.URI,track,Album.Name,Artist.Name.s.,Release.Date,Added.At,Record.Label,Genre, Key))

#recombine one-hot genre features and required numeric features
full_test <- cbind(sub_test,test_onehot)

row.names(full_test) <- NULL

#set a few features as factors
full_test$Explicit <- factor(full_test$Explicit)
full_test$Mode <- factor(full_test$Mode)
full_test$Time.Signature <- factor(full_test$Time.Signature)
full_test$Added.by <- factor(full_test$Added.by)

#manually specify the numeric features to be normalized
scaling_features_test <- c("Duration..ms.", "Popularity", "Danceability", "Energy", "Loudness", "Speechiness", "Acousticness", "Instrumentalness", "Liveness", "Valence", "Tempo")

#pull numeric features
numerics_test <- full_test[, scaling_features_test]

#scale numeric features
numerics_scaled_test <- as.data.frame(scale(numerics_test))

#remove unscaled features
full_sub_test <- full_test %>%
  select(-scaling_features_test)

#put scaled, categorical, and one-hot features together
cat_data_test <- cbind(full_sub_test,numerics_scaled_test)
row.names(cat_data_test) <- NULL

#ensure test data has all the same columns as training data
missing_cols <- setdiff(names(cat_data), names(cat_data_test))
if(length(missing_cols) > 0) {
  cat_data_test[missing_cols] <- 0
}
#ensure columns are in the same order as training data
cat_data_test <- cat_data_test[, names(cat_data)]
```

```{r}
#setting test data up for prediction
X_m1 <- model.matrix(Added.by ~ ., data = cat_data_test)[,-1]
y_test <- cat_data_test$Added.by

#create df for RF model
dtest <- data.frame(
  Matrix::as.matrix(X_m1),
  Added.by = factor(y_test)
)

#predict using sub_model 1 (multinomial logistic)
preds_prob_test <- predict(
  fit.lasso, 
  newx = X_m1, 
  s    = best_s_1, 
  type = "response"
)

#pull results and set up for stacking
prob_df_1_test <- as.data.frame(preds_prob_test)

k <- 2                    
names(prob_df_1_test) <- substr(names(prob_df_1_test), 1, nchar(names(prob_df_1_test)) - k)

#use fitted RF model to predict on test data
preds2_test <- predict(final_rf, dtest, type="prob")

#put rf results in proper form for stacking
k <- 6                       
colnames(preds2_test) <- substring(names(preds2_test), k + 1)

prob_df_2_test <- preds2_test
prob_df_2_test <- as.data.frame(prob_df_2_test)

#setting up stacked model data
colnames(prob_df_2_test)  <- levels(y_test)
colnames(prob_df_1_test) <- levels(y_test)

stack_df_test <- data.frame(
  prob_df_2_test,
  prob_df_1_test    
)
names(stack_df_test) <- paste0(rep(c("rf_", "glm_"), each = 10),
                          rep(head(levels(y_test), 10), 2))

#use stack model fit on training data to predict on the test data
final_model <- predict(
  stack_fit, 
  newx = as.matrix(stack_df_test), 
  s    = best_s, 
  type = "response"
)

#the final output, a n by 10 matrix giving the prob predicted for each response level for each observation in the test data
prob_mat_final_test <- final_model[,,1]

#setting up df for computing log loss of model on test data
loss_tbl <- as_tibble(prob_mat_final_test) %>%
  mutate(.truth = y_test, .rowid = row_number()) %>%
  relocate(.truth, .rowid)

#return log loss
mn_log_loss(loss_tbl, truth = .truth, !!!syms(colnames(prob_mat_final_test)))
```




